---
title: "Econometrics 2 capstone progress report code"
author: "Josh Copeland, Jocelyn Koswara and Ryan Luo"
date: "2024-09-08"
output: html_document
---

```{r setup_include_false, include=FALSE}
library(tidyverse)
library(naniar)
library(sjPlot)
```

```{r setup_include, include=FALSE}

#Packages used for this script:

# - tidyverse
# - naniar
# - sjPlot

```

# Importing and cleaning data

<br>

Tables used for the progress report:

- Psychology (S10AI)
- Housing: water (S12AI)
- Household background information (S1D)
- Key household information (key_hhld_info)

In order to derive the following variables:

- Binary variable indicating mental health status (1 = likely to have a mental health disorder) (S10AI)
- Binary variable indicating access to basic drinking water services (1 = has access) (S1D)
- Age (S10AI)
- Binary variable indicating sex (1 = female) (S10AI)
- Binary variable indicating religious minority (1 = not Christian) (S1D)
- Binary variable indicating if the person lives in an urban or rural area (1 = in an urban area) (S1D)

Analysis in this markdown document is separated by each data table imported.

### Importing the Pyschology table

```{r importing_data, message=FALSE, warning=FALSE}

################################################################################
################################ PSYCHOLOGY TABLE ##############################
################################################################################

s10ai <- read_csv("data/S10AI.csv") %>% 
  select(hhno, hhmid, depression, sex = s1d_1, age = s1d_4i) %>% 
  
  #Creating a new column as our depression_dummy. Kessler scores between 10-19 have a score of one in the data (== "likely to be well"). Anyone with scored higher than this has a score > 1, which classifies them as likely to have at least a mild disorder.
  mutate(depression_dummy = case_when(
    
    depression > 1 ~ 1, # Depressed
    TRUE ~ 0 # Not depressed
    
  )) %>% 
  
  # Turning sex into a dummy variable (1 == female)
  
  mutate(sex = case_when(
    
    sex == 1 ~ 0,
    sex == 2 ~ 1
    
  ))


######################## EXTRACTING JUST THE RELEVANT VARIABLES ################

s10ai <- s10ai %>% 
  select(hhno, hhmid, depression_dummy, sex_dummy = sex, age)



```

### Importing the housing table

<br>

We are importing this table to create a dummy variable for access to basic drinking services.

<br>

UNICEF defines a household's access to water as "basic" if it satisfies the following conditions:

 - It's delivered from one of the following sources: piped water, boreholes, tubewells, protected dug well, protected springs, rainwater and packaged of delivered water.
 
 - A round trip to collect water does not exceed 30 minutes.

``` {r importing_ii}

################################################################################
############################### HOUSING TABLES #################################
################################################################################

############################### WATER TABLE ####################################







s12ai <- read_csv("data/S12AI.csv") %>% 
  select(hhno,drinking_source = s12a_9i, drinking_source_distance_length = s12a_10ai, distance_unit = s12a_10aii, drinking_source_distance_mins = s12a_11) %>% 
  
  #Editing the drinking_source_distance cells to make them all the same scale: kilometres.
    
  mutate(drinking_source_distance_length = case_when(
    
    distance_unit == 0 ~ 0,  # In house
    distance_unit == 1 ~ as.numeric(drinking_source_distance_length) * 0.0009144,  # Yards to kilometers
    distance_unit == 2 ~ as.numeric(drinking_source_distance_length) / 1000,  # Meters to kilometers
    distance_unit == 3 ~ as.numeric(drinking_source_distance_length),  # Already in kilometers
    distance_unit == 4 ~ as.numeric(drinking_source_distance_length) * 1.609344,  # Miles to kilometers
    TRUE ~ drinking_source_distance_length
    
  ))
  


vis_miss(s12ai)

gg_miss_var(s12ai)


```

The charts above shows us that there is a lot of missing values for the distance variables in both length and mins. This likely have something todo with the drinking source of each household. I need to collect all the NA data together in order to diagnose the problem.

The charts below show us that:

 - Most of the problem is in 1 and 2, which correspond to plumbing in the house. We can change their distances to zero.
 
 - 8 is also a clear problem, which is bottled water. We think its reasonable to assume this botteld water is available at the house, so can change this distance to zero as well.
 
- 9 and 10 are protected wells and boreholes. Without more information about how far away they are (unavailable) we need to leave these as NAs.

``` {r importing_iii}

# Extracting and charting NA data

na_data <- s12ai %>% 
  filter(is.na(drinking_source_distance_length)) %>% 
  group_by(drinking_source) %>% 
  summarise(na_count = n())
  
ggplot(na_data, aes(x = drinking_source, y = na_count)) +
  geom_bar(stat = "identity") + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 14))




# Now I have diagnosed the problem, I need to make the necessary changes to the dataframe such that dirnking_sources with values 1 and 2 have a distance of zero in both length and minutes. All other NAs remain given data limitations.

s12ai <- s12ai %>% 
  mutate(drinking_source_distance_length = case_when(
    
    is.na(distance_unit) & drinking_source %in% c(1, 2, 8) ~ 0,
    TRUE ~ drinking_source_distance_length
  )) %>% 
  
  mutate(drinking_source_distance_mins = case_when(
        is.na(distance_unit) & drinking_source %in% c(1, 2, 8) ~ 0,
    TRUE ~ drinking_source_distance_mins
  ))



# Repeating the NA value analysis/chart below, the scale are now sufficiently small to continue/we don't have any other information that could help reduce the incidence of NAs.

na_data <- s12ai %>% 
  filter(is.na(drinking_source_distance_length)) %>% 
  group_by(drinking_source) %>% 
  summarise(na_count = n())

ggplot(na_data, aes(x = drinking_source, y = na_count)) +
  geom_bar(stat = "identity") + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 14))

gg_miss_var(s12ai)


# Because we can't deal with the remaining NAs, we exclude them from our analysis. However, we only exclude where NAs appear in the drinking_source_distance_length and drinking_source_distance_mins  variables.

s12ai <- s12ai %>% 
  filter(!is.na(drinking_source_distance_length)) %>% 
  filter(!is.na(drinking_source_distance_mins))

vis_miss(s12ai)


```

<br> 

Now we can actually produce our dummy variable for access to "basic drinking services".

```{r importing_iv}

s12ai <- s12ai %>% 
  mutate(basic_access_dummy = case_when(
    
    drinking_source_distance_mins <= 30 & 
      drinking_source %in% c(1, # Indoor plumbing
                             2, # Inside standpipe
                             5, # Pipe in niehgbouring household
                             6, # Private outside standpipe/tap
                             7, # Public standpipe
                             8, # Sachet/bottled water
                             9, # Borehole
                             10) # Protected well
    ~ 1,
    TRUE ~ 0
  ))


```




<br>

# Importing the hosehold background information table

<br>

```{r hh_background}

######################### RELIGIOUS MINORITY DUMMY #############################


s1d <- read_csv("data/S1D.csv") %>% 
  select(hhno, hhmid, religion = s1d_13, ethnicity = s1d_16) %>% 
  mutate(not_christian_dummy = 0) %>% 
  mutate(not_christian_dummy = case_when(
    
    # The following values of religion correspond with Christianity: 1,2,3,4,5 and 7.
    
    religion %in% c(1,2,3,4,5,7) ~ 0, 
    TRUE ~ 1
    
  ))

# Is it reasonable to think of Christian as the relgious majority? The chart below suggest they account for ~ 60% of the population. Therefore, it's reasonable to account for non-Christians are part of the relgious minority in Ghana.

religion_dummy_frequency <- s1d %>% 
  group_by(not_christian_dummy) %>% 
    summarise(count = n()) %>%
  mutate(percentage = (count / sum(count)) * 100) 
  
  

ggplot(religion_dummy_frequency, aes(not_christian_dummy, percentage)) + geom_bar(stat = "identity")




######################## EXTRACTING JUST THE RELEVANT VARIABLES ################

s1d <- s1d %>% 
  select(hhno, hhmid, not_christian_dummy)

```


# Importing key household information

```{r, key_hhld_info}

key_hhld_info <- read_csv("data/key_hhld_info.csv") %>%
  select(hhno, rural_dummy = urbrur) %>%
  mutate(rural_dummy = case_when(
    
    rural_dummy == "1" ~ 0,
    TRUE ~ 1
    
  ))


```

# Joining data

<br> 

Household data is not provided at the individual level. Therefore, we need to append it to our psychological data.

Doing a quick NA visualisation I can see that there are a few columns with NA values. Given how small they are as proportions, I omit the NA values for depression and drinking_source_distance. I don't both with distance_unit (its only use was to help us clean the data earlier.)


``` {r joining}

data <- s10ai %>% 
  inner_join(s12ai, by = "hhno") %>% 
  inner_join(key_hhld_info, by = "hhno") %>% 
  inner_join(s1d, by = c("hhno", "hhmid")) %>%  # This data is collected on the individual, therefore we need to join at the sub-household level.

  mutate(id = hhno + hhmid) %>%  # Creating a single hh identifier column
  
  select(id, depression_dummy, basic_access_dummy, drinking_source_distance_mins, drinking_source_distance_length, sex_dummy, age, not_christian_dummy, rural_dummy) #getting data columns into a helpful order
  
vis_miss(data)


# Omitting the very few remaining NA values

data <- data %>% 
  na.omit()

```
<br>

# Creating summary statistics

```{r summary_stats}


vars <- colnames(data)[!colnames(data) %in% c("id")]

# Create summary statistics
summary_stats <- data %>%
  summarise(across(all_of(vars), 
                   list(
                     mean = ~ mean(.x, na.rm = TRUE),
                     sd = ~ sd(.x, na.rm = TRUE),
                     min = ~ min(.x, na.rm = TRUE),
                     max = ~ max(.x, na.rm = TRUE)
                   ), 
                   .names = "{.col}_{.fn}")) 

# Reshape to long format
summary_stats <- summary_stats %>%
  pivot_longer(cols = everything(),  
               names_to = c("variable", "statistic"), 
               names_pattern = "(.*)_(.*)") %>%   # Match everything before the last underscore
  mutate(value = round(value,2))


summary_stats <- summary_stats %>%
  pivot_wider(names_from = statistic, values_from = value)

summary_stats$max <- format(summary_stats$max, scientific = FALSE)


print(summary_stats)


#################################### SAVING OFF DATA ###########################

write_csv(summary_stats, "summary_stats.csv")

write_csv(data, "data.csv")


```


# Producing linear and multiple regressions

```{r producing_models}

linear_model <- lm(depression_dummy ~ basic_access_dummy, data = data)

multiple_model <- lm(depression_dummy ~ basic_access_dummy + sex_dummy + age + not_christian_dummy + rural_dummy, data = data)


tab_model(linear_model, multiple_model,
          pred.labels = c("Intercept", "Access to basic drinking services dummy", "Sex dummy", "Age", "Religious minority dummy", "Rural dummy"),
          dv.labels = c("Linear regression model", "Multiple regression model"),
          p.style = "stars",
          digits = 3,
          file = "regression_table.doc")


```

