---
title: "ECOM90024 - FEB - Week 3 tutorial"
output: html_document
date: "2024-03-04"
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```

# 1.Using the housing starts data contained in the file ushstarts.csv, estimate a seasonal dummy model in which there are four seasons corresponding to the meteorological seasons in the northern hemisphere.


### A) compute and plot the seasonal factors

```{r q1a}

housing_starts <- read_csv("ushstarts.csv") %>% mutate(date = seq(as.Date("1946-01-01"), as.Date("1994-11-11"), by = "month"))

s <- 12

nyears <- ceiling(length(housing_starts$date)/s)

spring <- rep(c(0,0,1,1,1,0,0,0,0,0,0,0), nyears)
summer <- rep(c(0,0,0,0,0,1,1,1,0,0,0,0), nyears)
fall <- rep(c(0,0,0,0,0,0,0,0,1,1,1,0), nyears)
winter <- rep(c(1,1,0,0,0,0,0,0,0,0,0,1), nyears)

spring <- spring[1:length(housing_starts$date)]
summer <- summer[1:length(housing_starts$date)]
fall <- fall[1:length(housing_starts$date)]
winter <- winter[1:length(housing_starts$date)]


seasmod <- lm(formula = hstarts ~ 0 + spring + summer + fall + winter, data = housing_starts)

summary(seasmod)


seasonal.f <- coef(seasmod)

plot(seasonal.f, type = "l")

```


### B) Generate and plot forecasts (point, confidence intervals and prediction intervals) for the forecast horizon h = 25

```{r q1b}

#Defining variables to produce forecast

h = 24

nyears.f = h/s + 1

spring.h <- rep(c(0,0,1,1,1,0,0,0,0,0,0,0), nyears.f)
summer.h <- rep(c(0,0,0,0,0,1,1,1,0,0,0,0), nyears.f)
fall.h <- rep(c(0,0,0,0,0,0,0,0,1,1,1,0), nyears.f)
winter.h <- rep(c(1,1,0,0,0,0,0,0,0,0,0,1), nyears.f)


spring.h <- spring.h[12:36]
summer.h <- summer.h[12:36]
fall.h <- fall.h[12:36]
winter.h <- winter.h[12:36]

xh <- data.frame(spring.h, summer.h, fall.h, winter.h)

colnames(xh) <- c("spring", "summer", "fall", "winter")


# Producing forecasts

point.forecast <- predict(seasmod, newdata = xh)
point.forecast <- c(rep(NA,length(housing_starts$date)))

ci.forecast <- data.frame(predict(seasmod, newdata = xh, interval = "confidence"))


pi.forecast <- data.frame(predict(seasmod, newdata = xh, interval = "prediction"))


ci.forecast.lwr <- c(rep(NA,length(housing_starts$date)),ci.forecast$lwr)
ci.forecast.upr <- c(rep(NA,length(housing_starts$date)),ci.forecast$upr)
pi.forecast.lwr <- c(rep(NA,length(housing_starts$date)),pi.forecast$lwr)
pi.forecast.upr <- c(rep(NA,length(housing_starts$date)),pi.forecast$upr)


# Plot the forecast




```




# 2.	You are a research analyst for the Reserve Bank of Australia, monitoring the household consumption expenditures of Australian households. You are tasked with analysing a sample of quarterly Household Final Consumption Expenditure data from the September quarter of 1959 to the December quarter of 2018. 


### a) Import the original series located in column FM of sheet Data1 (series ID: A2302484C) and generate an appropriate plot of the data. Describe the primary visual features of the time series.    


```{r q2a}

cons <- read_excel("5206002_expenditure_volume_measures.xls", 
                   sheet = "Data1", 
                   col_names = "cons", 
                   range = "FM11: FM248")


```

### b.)	Using the decompose() function, generate and plot a seasonally adjusted version of the series (i.e., the original series with the seasonal component removed). Make sure to justify any choices that you have made performing the seasonal adjustment. 

```{r q2b}



```

### c.)	Compare the seasonally adjusted series that you have computed with the seasonally adjusted data that is contained in column CI of sheet Data1 (series ID: A2304081W). 

```{r q2c}



```

### d.)	Using the seasonally adjusted series that you have computed, estimate a quadratic trend model and use your model estimates to generate and plot the fitted trend. 

```{r q2d}



```


### Using your estimated model, generate and plot forecasts (point, confidence intervals and prediction intervals) for the forecast horizon h=8. For your interval forecasts, let (1-α)=0.8.   

```{r q2e}



```
