---
title: "Econometrics 2 capstone proposal code"
author: "Josh Copeland, Jocelyn Koswara and Ryan Luo"
date: "2024-08-11"
output: html_document
---

```{r setup_include_false, include=FALSE}
library(tidyverse)
library(naniar)
```

```{r setup_include, include=FALSE}

#Packages used for this script:

# - tidyverse

```

# Importing and cleaning data

<br>

Tables used for this proposal:

- Psychology (S10AI)
- Housing: water (S12AI)
- Housing: sewage facilities (S12AII)
- Household background information (S1D)

In order to derive the following variables:

- Binary variable indicating mental health status (1 = likely to have a mental health disorder) (S10AI)
- Distance from drinking water source (S12AI)
- Binary variable indicating exposure to open sewerage in house (1 = exposure) (S12AII)
- Age (S10AI)
- Binary variable indicating sex (1 = female) (S10AI)
- Binary variable indicating religious minority (1 = not Christian) (S1D)
- Binary variable indicating ethnic minority (1 = not in an ethnicity that accounts for at least 5% of people surveyed) (S1D)


Analysis in this markdown document is separated by each data table imported.

### Importing the Pyschology table

```{r importing_data, message=FALSE, warning=FALSE}

################################################################################
################################ PSYCHOLOGY TABLE ##############################
################################################################################

s10ai <- read_csv("data/S10AI.csv") %>% 
  select(hhno, hhmid, depression, sex = s1d_1, age = s1d_4i) %>% 
  
  #Creating a new column as our depression_dummy. Kessler scores between 10-19 have a score of one in the data (== "likely to be well"). Anyone with scored higher than this has a score > 1, which classifies them as likely to have at least a mild disorder.
  mutate(depression_dummy = case_when(
    
    depression > 1 ~ 1, # Depressed
    TRUE ~ 0 # Not depressed
    
  )) %>% 
  
  # Turning sex into a dummy variable (1 == female)
  
  mutate(sex = case_when(
    
    sex == 1 ~ 0,
    sex == 2 ~ 1
    
  ))


######################## EXTRACTING JUST THE RELEVANT VARIABLES ################

s10ai <- s10ai %>% 
  select(hhno, hhmid, depression_dummy, sex_dummy = sex, age)



```

### Importing the housing tables

<br>

In the following order:

 - Water
 - Sewerage

``` {r importing_ii}

################################################################################
############################### HOUSING TABLES #################################
################################################################################

############################### WATER TABLE ####################################

s12ai <- read_csv("data/S12AI.csv") %>% 
  select(hhno,drinking_source = s12a_9i, drinking_source_distance = s12a_10ai, distance_unit = s12a_10aii) %>% 
  
  #Editing the drinking_source_distance cells to make them all the same scale: metres.
  
  mutate(drinking_source_distance = case_when(
    
    distance_unit == 0 ~ 0,  # In house
    distance_unit == 1 ~ 0,  # In yard (assuming 0 meters)
    distance_unit == 2 ~ as.numeric(drinking_source_distance),  # Already in meters
    distance_unit == 3 ~ as.numeric(drinking_source_distance) * 1000,  # Kilometers to meters
    distance_unit == 4 ~ as.numeric(drinking_source_distance) * 1609.344,  # Miles to meters
    TRUE ~ drinking_source_distance
    
  ))




vis_miss(s12ai)

gg_miss_var(s12ai)


```

<br> 

The chart above shows us that there is a lot of missing values for distance_unit. This likely have something to do with the drinking source of each household. I need to collect all the NA data together in order to diagnose the problem.

The chart below shows us that:

 - Most of the problem is in 1 and 2, which corresponds to plumbing in the house. We can change their distances to zero.

 - 8 is also a clear problem, which is bottled water. Therefore leave this as NA.
 
 - 9 and 10 are protected wells and boreholes. Without more information about how far away they are (unavailable) we need to leave these as NAs.

``` {r importing_iii}

# Extracting and charting NA data

na_data <- s12ai %>% 
  filter(is.na(drinking_source_distance)) %>% 
  group_by(drinking_source) %>% 
  summarise(na_count = n())
  
ggplot(na_data, aes(x = drinking_source, y = na_count)) +
  geom_bar(stat = "identity") + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 14))




# Now I have diagnosed the problem, I need to make the necessary changes to the dataframe such that dirnking_sources with values 1 and 2 have a distance of zero. All other NAs remain given data limitations.

s12ai <- s12ai %>% 
  mutate(drinking_source_distance = case_when(
    
    is.na(distance_unit) & drinking_source %in% c(1, 2) ~ 0,
    TRUE ~ drinking_source_distance
  ))



# Repeating the NA value analysis/chart below, the scale are now sufficiently small to continue/we don't have any other information that could help reduce the incidence of NAs.

na_data <- s12ai %>% 
  filter(is.na(drinking_source_distance)) %>% 
  group_by(drinking_source) %>% 
  summarise(na_count = n())

ggplot(na_data, aes(x = drinking_source, y = na_count)) +
  geom_bar(stat = "identity") + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 14))

gg_miss_var(s12ai)


# Because we can't deal with the remaining NAs, we exclude them from our analysis. However, we only exclude where NAs appear in the drinking_source_distance variable.

s12ai <- s12ai %>% 
  filter(!is.na(drinking_source_distance))

vis_miss(s12ai)

######################## EXTRACTING JUST THE RELEVANT VARIABLES ################

s12ai <- s12ai %>% 
  select(hhno, drinking_source_distance)


############################### SEWERAGE TABLE #################################

s12aii <- read_csv("data/S12AII.csv") %>% 
  select(hhno, sewerage_dummy = s12b_23)  
  # Is there any open sewer,drain in/around the house? Note: 1 == Yes, 2 == No, 3 == Drains are covered
  
frequency_test <- s12aii %>% 
    group_by(sewerage_dummy) %>% 
  summarise(frequency = n())

ggplot(frequency_test, aes(x = sewerage_dummy, y = frequency)) + geom_bar(stat = "identity") 


# Given how infrequently option 3 occurs, we are going to exclude it to ensure our variable actually as a dummy. We then need to make sure values are between 0 and 1.

s12aii <- s12aii %>% 
  filter(sewerage_dummy != 3) %>% 
  
  # Turning variable into actual dummy variable (exposure == 1)
  mutate(sewerage_exposure_dummy = case_when(
    
    sewerage_dummy == 2 ~ 0,
    sewerage_dummy == 1 ~ 1
    
  ))


######################## EXTRACTING JUST THE RELEVANT VARIABLES ################

s12aii <- s12aii %>% 
  select(hhno, sewerage_exposure_dummy)


```


<br>

# Importing the hosehold background information table

<br>

```{r hh_background}

######################### RELIGIOUS MINORITY DUMMY #############################


s1d <- read_csv("data/S1D.csv") %>% 
  select(hhno, hhmid, religion = s1d_13, ethnicity = s1d_16) %>% 
  mutate(not_christian_dummy = 0) %>% 
  mutate(not_christian_dummy = case_when(
    
    # The following values of religion correspond with Christianity: 1,2,3,4,5 and 7.
    
    religion %in% c(1,2,3,4,5,7) ~ 0, 
    TRUE ~ 1
    
  ))

# Is it reasonable to think of Christian as the relgious majority? The chart below suggest they account for ~ 60% of the population. Therefore, it's reasonable to account for non-Christians are part of the relgious minority in Ghana.

religion_dummy_frequency <- s1d %>% 
  group_by(not_christian_dummy) %>% 
    summarise(count = n()) %>%
  mutate(percentage = (count / sum(count)) * 100) 
  
  

ggplot(religion_dummy_frequency, aes(not_christian_dummy, percentage)) + geom_bar(stat = "identity")


########################### ETHNIC MINORITY DUMMY ##############################

# Making a call on ethnic minorities in Ghana is a bit more difficult. From the bar chart below, 5 minorities reflect at least 5 per cent of the population. We therefore define being in the ethnic minority as not being in these 5 ethnic groups. Those ethnicities in the majority are shown below.

#These groups in the majority are following: Asante, Ewe, Fante, Dagomba, Dagarte/Lobi, Kokomba.

ethnicity_analysis <- s1d %>% 
  group_by(ethnicity) %>% 
  summarise(count = n()) %>% 
  mutate(percentage = (count / sum(count)) * 100) %>% 
  arrange(desc(percentage))
  
ggplot(ethnicity_analysis, aes(ethnicity, percentage)) + geom_bar(stat = "identity")

print(head(ethnicity_analysis))


s1d <- s1d %>% 
  mutate(ethnic_minority_dummy = case_when(
    
    ethnicity %in% c(8,30,14,63,62,53) ~ 0,
    TRUE ~ 1
    
  ))


######################## EXTRACTING JUST THE RELEVANT VARIABLES ################

s1d <- s1d %>% 
  select(hhno, hhmid, not_christian_dummy, ethnic_minority_dummy)

```


# Joining data

<br> 

Household data is not provided at the individual level. Therefore, we need to append it to our psychological data.

Doing a quick NA visualisation I can see that there are a few columns with NA values. Given how small they are as proportions, I omit the NA values for depression and drinking_source_distance. I don't both with distance_unit (its only use was to help us clean the data earlier.)


``` {r joining}

data <- s10ai %>% 
  inner_join(s12ai, by = "hhno") %>% 
  inner_join(s12aii, by = "hhno") %>% 
  inner_join(s1d, by = c("hhno", "hhmid")) %>%  # This data is collected on the individual, therefore we need to join at the sub-household level.

  mutate(id = hhno + hhmid) %>%  # Creating a single hh identifier column
  
  select(id, depression_dummy, drinking_source_distance, sewerage_exposure_dummy, sex_dummy, age, not_christian_dummy, ethnic_minority_dummy) #getting data columns into a helpful order
  
vis_miss(data)


# Omitting the very few remaining NA values

data <- data %>% 
  na.omit()

```
<br>

# Creating summary statistics

```{r summary_stats}


vars <- colnames(data)[!colnames(data) %in% c("id")]

# Create summary statistics
summary_stats <- data %>%
  summarise(across(all_of(vars), 
                   list(
                     mean = ~ mean(.x, na.rm = TRUE),
                     sd = ~ sd(.x, na.rm = TRUE),
                     min = ~ min(.x, na.rm = TRUE),
                     max = ~ max(.x, na.rm = TRUE)
                   ), 
                   .names = "{.col}_{.fn}")) 

# Reshape to long format
summary_stats <- summary_stats %>%
  pivot_longer(cols = everything(),  
               names_to = c("variable", "statistic"), 
               names_pattern = "(.*)_(.*)") %>%   # Match everything before the last underscore
  mutate(value = round(value,2))


summary_stats <- summary_stats %>%
  pivot_wider(names_from = statistic, values_from = value)

summary_stats$max <- format(summary_stats$max, scientific = FALSE)


print(summary_stats)


#################################### SAVING OFF DATA ###########################

write_csv(summary_stats, "summary_stats.csv")

write_csv(data, "data.csv")


```

