---
title: "Econometrics 2 capstone final report models and analysis"
author: "Josh Copeland, Jocelyn Koswara and Ryan Luo"
date: "2024-10-1"
output: html_document
---

```{r setup_include_false, include=FALSE}
library(tidyverse)
library(AER)
library(stargazer)
library(lmtest)
library(sandwich)
library(rvest)
library(writexl)
library(openxlsx)
```

```{r setup_include, include=FALSE}

#Packages used for this script:

# - tidyverse
# - AER
# - stargazer

```

```{r data_import_models}

data <- read_csv("model_data.csv")

```


# Key models/output

```{r models}

#################################### MODELS ####################################

# linear regression model
m1 <- lm(depression_dummy ~ basic_access_dummy, data = data)


# Multiple linear regresion model
m2 <- lm(depression_dummy ~ basic_access_dummy + sex_dummy + age + not_christian_dummy + rural_dummy, data = data)


# First stage 2SLS model
m3 <- lm(basic_access_dummy ~ drinking_source_distance_mins + sex_dummy + age + not_christian_dummy + rural_dummy,
                  data = data)

# Second stage 2SLS model
m4 <- ivreg(depression_dummy ~ basic_access_dummy + sex_dummy + age + not_christian_dummy + rural_dummy |
      sex_dummy + age + not_christian_dummy + rural_dummy + drinking_source_distance_mins, data = data)


```
# Formatting output

```{r formatting_output}

#################################### OUTPUT TABLES ##############################


key_results <- stargazer(m1,
                         m2,
                         m3,
                         m4, 
                         type = "html",
                         title = "Table 2: Model estimates",
                           se = list(
                          sqrt(diag(vcovHC(m1))),
                          sqrt(diag(vcovHC(m2))),
                          sqrt(diag(vcovHC(m3))),
                          sqrt(diag(vcovHC(m4)))
                          ),
                        out = "Key_Results.html"  
                                
  )


##################### AUTOMATING OUTPUTTING TABLES TO XLSX FILE ################

key_results_table <- read_html("Key_Results.html") %>% 
    html_table(fill = TRUE) %>% 
    pluck(1) %>% 
    filter(if_any(c(X1, X2, X3), ~ . != ""))

write_xlsx(key_results_table, "key_results_table.xlsx")

```



# Robustness checks

## Changing definitions of water access

Ultimately, our chosen definition of water access has been chosen arbitrarily from UNICEF's taxonomy of access. If our results are robust, the size of our estimators should decrease as we move from the lowest ("no service") to highest ("safely managed") level of water access. The better the access, the greater an impact it should have on reducing the incidence of mental health issues.

```{r rc_water_check}

################################## CREATING MODELS #############################


#No service
m5 <- ivreg(depression_dummy ~ no_service_dummy + sex_dummy + age + not_christian_dummy + rural_dummy |
      sex_dummy + age + not_christian_dummy + rural_dummy + drinking_source_distance_mins, data = data)


# Unimproved
m6 <- ivreg(depression_dummy ~ unimproved_dummy + sex_dummy + age + not_christian_dummy + rural_dummy |
      sex_dummy + age + not_christian_dummy + rural_dummy + drinking_source_distance_mins, data = data)



# Limited
m7 <- ivreg(depression_dummy ~ limited_dummy + sex_dummy + age + not_christian_dummy + rural_dummy |
      sex_dummy + age + not_christian_dummy + rural_dummy + drinking_source_distance_mins, data = data)


# Basic access
m8 <- ivreg(depression_dummy ~ basic_access_dummy + sex_dummy + age + not_christian_dummy + rural_dummy |
      sex_dummy + age + not_christian_dummy + rural_dummy + drinking_source_distance_mins, data = data)


# Safely managed
m9 <- ivreg(depression_dummy ~ safely_managed_dummy + sex_dummy + age + not_christian_dummy + rural_dummy |
      sex_dummy + age + not_christian_dummy + rural_dummy + drinking_source_distance_mins, data = data)



################################# COMPARING OUTPUT #############################

rc_water_def_results <- stargazer(
          m5,
          m6,
          m7,
          m8,
          m9,
          type = "html",
          title = "Table X: Changing water definition estimates",
          se = list(
          sqrt(diag(vcovHC(m5))),  
          sqrt(diag(vcovHC(m6))),
          sqrt(diag(vcovHC(m7))),
          sqrt(diag(vcovHC(m8))),
          sqrt(diag(vcovHC(m9)))
          ),
          out = "RC_Water_Def_Results.html"
          )

##################### AUTOMATING OUTPUTTING TABLES TO XLSX FILE ################

rc_water_def_results_table <- read_html("RC_Water_Def_Results.html") %>% 
    html_table(fill = TRUE) %>% 
    pluck(1) %>% 
    filter(if_any(c(X1, X2, X3, X4,X5, X6), ~ . != ""))

write_xlsx(rc_water_def_results_table, "rc_water_def_results_table.xlsx")




##################### AUTOMATING OUTPUTTING TABLES TO XLSX FILE ################

key_results_table <- read_html("Key_Results.html") %>% 
    html_table(fill = TRUE) %>% 
    pluck(1) %>% 
    filter(if_any(c(X1, X2, X3), ~ . != ""))

write_xlsx(key_results_table, "key_results_table.xlsx")


```